---
title: "Midterm Report of STAT3621 Statistical Data Analysis"
author: "Lau Yan Chun Chris"
date: "`r Sys.Date()`"
output: pdf_document
---

# Environment Set Up
```{r setup, echo = T}
setwd("/Users/ycclau/Desktop/Academic/STAT3621/Assignment/A2")

libraries <- c("mlbench","ggpubr","corrplot", "viridis", "leaps", "glmnet","car")

importLibrary <- function(library_name) {
  if (!require(library_name, character.only = TRUE)) {
    install.packages(library_name)
    library(library_name, character.only = TRUE)
  }
}

for(library_name in libraries){
  importLibrary(library_name)
}

if (!require(tidyverse)) { install.packages("tidyverse")
  library(ggplot2)
  library(dplyr)
}
```
# Data loading & Checking for NA, Null value
```{r echo = T}
data("BostonHousing")
summary(BostonHousing)

data <- BostonHousing

#Check NA & Null value
cat("dimension of the dataset:",dim(data),"\n")
cat("Number of NA in dataset:",sum(is.na(data)),"\nNumber of Null in dataset: ",sum(is.null(data)))
```

# Understanding the Data 

##Univarate EDA
```{r echo = T}
var.names <- names(data)
univarate.plots.list <- list()

for (i in 1:length(var.names)) {
  plt <- ggplot(data = data, aes(x = .data[[var.names[i]]]))
  
  # Check if the variable is discrete
  if (is.factor(data[[var.names[i]]])) {
    plt <- plt + geom_bar(fill = "skyblue", stat = "count", width = 0.5, alpha = 0.8)
  } else {
    plt <- plt + geom_density(color = "darkblue", fill = "skyblue", alpha = 0.5)
  }
  
  univarate.plots.list <- c(univarate.plots.list, list(plt))
}

univarate.plots <- ggarrange(plotlist = univarate.plots.list, 
                             ncol = 4, nrow = 4)

annotate_figure(univarate.plots, 
                top = text_grob("Univariate Analysis for BostonHousing dataset", 
                size = 14))
```

## Correlation for non-binary variables
```{r echo = T}
data.wo.chas <- as.data.frame(data)
data.wo.chas[["chas"]] <- NULL

data.wo.chas.cor = cor(data.wo.chas)
corrplot(data.wo.chas.cor, 
         method = "ellipse", 
         title = "Correlation table for BostonHousing (w/out chas)",
         mar = c(0,0,1,0),
         order = 'hclust'
         )
```

## Relationship of variables and binary variable ```chas```

###Distribution of ```chas```
```{r echo = T}
table(data$chas)
```

### Boxplots for variables versus ```chas```
```{r echo = T}
chas.boxplots.list = list()

for (i in 1:length(var.names)){
  if (var.names[i] == "chas"){next}
  
  plt <- ggplot(data, aes(x = chas, y = .data[[var.names[i]]],fill = chas)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.8) +
    theme(legend.position="none") +
    scale_color_brewer(palette="BuPu")
  
  chas.boxplots.list <- c(chas.boxplots.list, list(plt))
}

chas.boxplots <- ggarrange(plotlist = chas.boxplots.list, 
                          ncol = 7, nrow = 2)

annotate_figure(chas.boxplots, 
                top = text_grob("Boxplot for variables by Factor chas", 
                size = 14))
```

### Wilcoxon Rank Sum Test
```{r echo = T}
print("Wilcoxon Rank Sum Test for variables by chas")
chas.wilcox.results = list()
chas.diff.list = list()
for (i in 1:length(var.names)){
  if(var.names[i] == "chas"){next}
  
  control <- "chas"
  target <- as.formula(paste(var.names[i], "~", control))
  
  result <- wilcox.test(target, data = data)
  
  title <- result$data.name
  W <- result$statistic
  p <- result$p.value
  H0_accepted <- TRUE
  if( p < 0.05){
    H0_accepted <- FALSE
    chas.diff.list <- c(chas.diff.list, list(var.names[[i]]))
  }
  cat(title,"\n","W:",W,"\t", "p-value:",p,"\t","H0 accepted?(No diff.)", H0_accepted, "\n")
}
```
####Desity plot for variables that ```chas``` matters
```{r echo = T}

chas.diff.plts.list <- list()
for(i in 1:length(chas.diff.list)){
  var_name <- chas.diff.list[[i]]
  plt <- ggplot(data, aes(x = .data[[var_name]], color = chas, fill = chas)) +
    geom_density(alpha = 0.3)
    chas.diff.plts.list <- c(chas.diff.plts.list, list(plt))
}

chas.diff.plts <- ggarrange(plotlist = chas.diff.plts.list, 
                             ncol = 1, nrow = 3)
annotate_figure(chas.diff.plts, 
                top = text_grob("Density Comparison of Variables by chas Factor", 
                size = 14))
```

# Statement of the problem
1. Prediction of ```medv``` using linear regression model

## variable distributions versus ```meddv``` without outliers
```{r echo = T}
medv.scatter.list <- list()

for (i in 1:length(var.names)){
  if (var.names[i] == "chas" || var.names[i] == "medv"){next}
  
  formula <- as.formula(paste("medv ~", var.names[i]))
  model <- lm(formula, data = data)
  cooksd <- cooks.distance(model)
  cooksd_threshold <- 4 / length(cooksd)
  outliers <- which(cooksd > cooksd_threshold)
  data_focus <- data[-outliers,]

  
  plt <- ggplot(data_focus, aes(x = .data[[var.names[i]]], y = medv)) +
    geom_point(color = "aquamarine3", alpha = 0.8, size = 0.5) +
    geom_point(color = "cyan4", alpha = 0.8, size = 0.2) + 
    geom_smooth(method = "lm", se = FALSE, color = "skyblue", alpha = 0.8)
    
  medv.scatter.list <- c(medv.scatter.list, list(plt))
}

medv.scatter <- ggarrange(plotlist = medv.scatter.list, 
                          ncol = 4, nrow = 3)
annotate_figure(medv.scatter, top = "SLR for variables versus medv")
```

```{r echo = T}

ResidualPlot <- function(model, xlab, title = ""){
  plt <-ggplot(model, aes(x= .fitted, y = .resid)) +
      geom_point(color = "darkblue", alpha = 0.8, size = 0.5) +
      geom_point(color = "skyblue", alpha = 0.8, size = 0.2) +
      geom_hline(yintercept = 0) +
      geom_smooth(method = "loess", alpha = 0.5, color = "black") +
      xlab(xlab)
    
  return(plt)
}

medv.resid.list <- list()

for (name in var.names){
  if (name == "chas" || name == "medv"){next}
  
  formula <- as.formula(paste("medv ~", name))
  model <- lm(formula, data = data)
  plt <- ResidualPlot(model, xlab = paste("~",name))
  medv.resid.list <- c(medv.resid.list, list(plt))
}

medv.resid <- ggarrange(plotlist = medv.resid.list, 
                          ncol = 4, nrow = 3)

annotate_figure(medv.resid, top = "Residual Plot for variables")
```
### Log-transformation
```{r echo = T}

logResidualPlot <- function(name, data){
  df <- data.frame(data$medv, sapply(data[[name]],log10))
  colnames(df)[1] <- "medv"
  model <- lm(medv ~., data = df)
  plt <- ResidualPlot(model, xlab = paste("~log_",name))
  return(plt)
}
crim.log <- logResidualPlot("crim", data)
rad.log <- logResidualPlot("rad", data)
tax.log <- logResidualPlot("tax", data)


log.list <- ggarrange(plotlist = list(crim.log, rad.log, tax.log), 
                          ncol = 3, nrow = 3)

annotate_figure(log.list, top = "Residual Plot for log(features)")

```

### Polynomial Approach
```{r echo = T}

showPolyPlot <- function(name, data){
  df.x1 <- data.frame(medv = data$medv, data[[name]])
  df.x2 <- data.frame(medv =data$medv, data[[name]], 
                      sapply(data[[name]],function(x) x**2))
  df.x3 <- data.frame(medv =data$medv, data[[name]], 
                      sapply(data[[name]],function(x) x**2),
                      sapply(data[[name]],function(x) x**3))
  df.x4 <- data.frame(medv =data$medv, data[[name]], 
                      sapply(data[[name]],function(x) x**2),
                      sapply(data[[name]],function(x) x**3),
                      sapply(data[[name]],function(x) x**4))
  df.x5 <- data.frame(medv =data$medv, data[[name]], 
                      sapply(data[[name]],function(x) x**2),
                      sapply(data[[name]],function(x) x**3),
                      sapply(data[[name]],function(x) x**4),
                      sapply(data[[name]],function(x) x**5))
  df.x6 <- data.frame(medv =data$medv, data[[name]], 
                      sapply(data[[name]],function(x) x**2),
                      sapply(data[[name]],function(x) x**3),
                      sapply(data[[name]],function(x) x**4),
                      sapply(data[[name]],function(x) x**5),
                      sapply(data[[name]],function(x) x**6))
  
  df.xList <- list(df.x1, df.x2, df.x3, df.x4, df.x5, df.x6)
  plt.list <- list()
  i = 1
  for(xn in df.xList){
    model <- lm(medv ~., data = xn)

    plt <- ResidualPlot(model, xlab = paste0("T_",i,"(",name,")"))
    plt.list <- c(plt.list, list(plt))
    i <- i + 1
  }
  
  plt.group <- ggarrange(plotlist = plt.list, 
                          ncol = 3, nrow = 2)

  annotate_figure(plt.group, top = paste0("Residual Plot for Poly.T {1to6} of ",name))
}
```

```{r echo = T}
showPolyPlot("crim", data)
showPolyPlot("rm", data)
showPolyPlot("ptratio", data)
showPolyPlot("lstat", data)
```
## Analysis method
```{r echo = T}

performSubsetSelection <- function(data, method, nmax = 13) {
  subset <- regsubsets(medv ~ ., data = data, method = method, nvmax = nmax)
  
  # Print the method name
  cat("Method:", method, "\n")
  
  # Print summary statistics
  summary <- summary(subset)
  cat("R2_adj:\n")
  print(summary$adjr2)
  cat("max. R2_adj: ",max(summary$adjr2), "Model suggested:",which.max(summary$adjr2),"\n")
  
  cat("Mallows' Cp:\n")
  print(summary$cp)
  cat("Diff. of measured and theoretical value:\n")
  Cp.theorical <- 1:nmax
  Cp.diff <- summary$cp - Cp.theorical
  print(round(Cp.diff, 5))
  cat("min. diff. Cp: ",min(Cp.diff), "Model suggested:",which.min(Cp.diff),"\n")
  
  cat("BIC:\n")
  print(summary$bic)
  cat("min. BIC: ",min(summary$bic), "Model suggested:",which.min(summary$bic),"\n")

  return(summary)
}

models.ex <- performSubsetSelection(data, "exhaustive")
```

## best model without any transformation with exhaustive method
```{r echo = T}
##change the position of binary list since medv locates at the end in dataset
##only for the original data frame
model.best <- models.ex$which[11, ]
model.best <- model.best[-1]
model.best <- c(model.best, TRUE)

model.best.data <- data[model.best]

model.m1 <- lm(medv ~., data =model.best.data)
summary(model.m1)
```

## Residual Analysis
```{r echo = T}
ResidualPlot(model.m1, xlab = "X_m1(p=11)")
```
## Dealing with non-linearity
```{r echo =T}
crim.x2 <-sapply(data$crim, function(x,n) x**n, 2)
crim.x3 <-sapply(data$crim, function(x,n) x**n, 3)
rm.x2 <- sapply(data$rm, function(x,n) x**n, 2)
rm.x3 <- sapply(data$rm, function(x,n) x**n, 3)
rad.log <- sapply(data$rad, log10)
tax.log <- sapply(data$tax, log10)
ptratio.x2 <- sapply(data$ptratio, function(x,n) x**n, 2)
ptratio.x3 <- sapply(data$ptratio, function(x,n) x**n, 3)
ptratio.x4 <- sapply(data$ptratio, function(x,n) x**n, 4)
lstat.x2 <- sapply(data$lstat, function(x,n) x**n, 2)
lstat.x3 <- sapply(data$lstat, function(x,n) x**n, 3)
lstat.x4 <- sapply(data$lstat, function(x,n) x**n, 4)

data.trans <- data.frame(medv = data$medv,
                         crim = data$crim, crim.x2, crim.x3,
                         zn = data$zn,
                         chas = data$chas,
                         nox = data$nox,
                         rm = data$rm, rm.x2, rm.x3,
                         dis = data$dis,
                         rad.log,
                         tax.log,
                         ptratio = data$ptratio, ptratio.x2, ptratio.x3, ptratio.x4,
                         b = data$b,
                         lstat = data$lstat, lstat.x2, lstat.x3, lstat.x4)

print(dim(data.trans))
```
```{r echo = T}
models.trans <- performSubsetSelection(data.trans, "exhaustive", nmax = 20)
```


## Multicollinearity
```{r echo = T}
data.trans.m1 <- data.trans[models.trans$which[17,]]
model.trans.m1 <- lm(medv ~ ., data = data.trans.m1)

vif <- 1 / vif(model.trans.m1)

threshold <- 5

if(any(vif) > threshold){
  print("Multicollinearity attention is raised.")
} else{
  cat("No multicollinearity is greater than threshold.\n")
}

print(vif)
```

```{r echo = T}
model.trans.m1$coefficients
ResidualPlot(model.trans.m1, xlab = "T_{17}(X)")
```
### Principal components analysis (PCA) for data visualization
```{r echo = T}
predictors <- model.matrix(model.trans.m1)[,-1]
predictors.wo.b0 <- predictors[,-1]
pca.result <- prcomp(predictors.wo.b0, scale. = TRUE)
pca.scaled.vec <- as.data.frame(predict(pca.result, newdata = predictors)[, 1:2])

pca.scaled.vec$medv_pca <- data$medv

ggplot(pca.scaled.vec, aes(x = PC1, y = PC2, color = medv_pca)) +
  geom_point() +
  ggtitle(" medv model visalized by PCA (dim = 2) ")
```





